---
title: "script_2.3a_analyse_pbmcs_20250805"
author: "GKM"
date: "2025-08-18"
output:
  html_document:
    keep_md: TRUE
---

In script 2.1 and 2.2a, the 20250508 data was loaded, HTOs demultimplexed, filtered for QC, then dimensional reduction and clustering was performed. The processed seurat object was saved as were the cluster biomarkers.

These can now be read in to avoid repeating the prior steps.

```{r}
library("dplyr")
library("Seurat")
library("ggplot2")
library("readr")
library("tidyr")
library("patchwork")
```

# Create file path for saving plots
```{r}
#file_path <- "/scratch/prj/id_hill_sims_wellcda/scCHANTS/20250805_run1/"
file_path <- "/Users/k2477939/Library/CloudStorage/OneDrive-King\'sCollegeLondon/General\ -\ Laboratory\ of\ Molecular\ Immunotherapy\ and\ Antibiotics/Data/Georgia\ Miller/scCHANTS/"

plot_path <- paste0(file_path, "20250805_run1/plots/PBMC_plots/")
```

```{r}
# load seurat object
scCHANTS_pbmc <- readRDS(paste0(file_path, "20250805_run1/20250805_scCHANTS_pbmc_processed.Rds"))

# check has pca and umap saved
names(scCHANTS_pbmc@reductions)

# check has RNA_snn_res column with cell_types
colnames(scCHANTS_pbmc@meta.data)

Idents(scCHANTS_pbmc) <- "seurat_clusters"

metadata <- scCHANTS_pbmc@meta.data

# load in cluster markers
pbmc_markers <- read_csv(paste0(file_path, "20250805_run1/20250805_scCHANTS_pbmc_markers.csv"))

head(pbmc_markers)

```

# Look at cluster markers


can also: Classification power, ROC test while running findmarkers returns value 0-1 (random-perfect classification):
cluster0.markers <- FindMarkers(scCHANTS_pbmc, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
    
```{r fig.dim = c(10,15)}
pbmc_markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>% 
  arrange(avg_log2FC) %>% 
  head()

top10 <- pbmc_markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>% # take top 10 genes for each cluster
    ungroup()
top10

c1 <- DoHeatmap(scCHANTS_pbmc, features = top10$gene) 
  NoLegend()
c1
# ggsave("top10_markers_heatmap.png", plot = c1, device = png, width = 17, height = 25, unit = "cm", path = paste0(file_path, "/cell_types/") )
```

```{r}
c2 <- DimPlot(scCHANTS_pbmc, group.by = "seurat_clusters", label = TRUE)
c2
# ggsave("by_clusters_umap.png", plot = c2, device = png, width = 15, height = 10, unit = "cm", path = paste0(file_path, "/cell_types/") )
```

```{r fig.dim = c(15,15)}
# 1 marker for each cluster
c3 <- FeaturePlot(scCHANTS_pbmc, features = c("ITK", "IL1B", "CD74", "CRIP1", "GNLY", "CD74", "SELL", "RGCC", "CD8A", "TXN", "MT1X", "LINC02211"))
c3
# ggsave("top_marker_featureplots.png", plot = c3, device = png, width = 25, height = 20, unit = "cm", path = paste0(file_path, "/cell_types/") )

c4 <- VlnPlot(scCHANTS_pbmc, features = c("ITK", "IL1B", "CD74", "CRIP1", "GNLY", "CD74", "SELL", "RGCC", "CD8A", "TXN", "MT1X", "LINC02211"))
c4
# ggsave("top_marker_vlnplots.png", plot = c4, device = png, width = 25, height = 20, unit = "cm", path = paste0(file_path, "/cell_types/") )

```

# Load known broad cell type markers in, make dotplot

```{r fig.dim = c(12,10)}
# plot dotplot of expected cell types for markers
# read in binary tables for specific and broad cell types
markers_binary_sp <- read_csv(paste0(file_path, "marker_genes/marker_genes_binary_specificcelltypes.csv"))

markers_binary_br <- read_csv(paste0(file_path, "marker_genes/marker_genes_binary_broadcelltypes.csv"))
         
# has only broad cell types but all gene markers so can compare directly
markers_binary_br2 <- read_csv(paste0(file_path, "marker_genes/marker_genes_binary_broadcelltypes_allmarkers.csv"))

# make into long format so can plot dotplot
markers_long_sp <- markers_binary_sp %>%
  pivot_longer(
    cols = -marker,
    names_to = "cell_type",
    values_to = "value"
  ) %>% 
  mutate(marker = factor(marker,
                         levels = c("CD3D", "CD3E", "CD3G", "CCR7", "SELL", "CD8A", "CD8B", "LTB", "IL7R", "S100A4", "GZMB", "PRF1", "FCGR3A", "NKG7", "GNLY", "KLRC1", "CD14", "LYZ", "S100A9", "CSF3R", "CSF1R", "LYN", "IFITM1", "IFITM2", "IFITM3", "CD79A", "MS4A1", "CD1C", "ITGAX", "CLEC4C", "GP9", "ITGA2B", "PF4", "PPBP")),
          cell_type = factor(cell_type,
                             levels = c("CD4_naive", "CD4_memory", "CD8_naive", "CD8_memory", "NK_dim", "NK_bright", "mono_classic", "mono_nonclassic", "B_normal", "B_plasma", "DC_myeloid", "DC_plasmacytoid", "megak")))

markers_long_br <- markers_binary_br %>%
  pivot_longer(
    cols = -marker,
    names_to = "cell_type",
    values_to = "value"
  ) %>% 
  mutate(marker = factor(marker,
                         levels = c("CD3D", "CD3E", "CD3G", "CD8A", "CD8B", "GZMB", "PRF1", "CD14", "CD79A", "GP9", "ITGA2B", "PF4", "PPBP")),
         cell_type = factor(cell_type,
                             levels = c("CD4T", "CD8T", "NK", "mono", "B", "DC", "megak")))

markers_long_br2 <- markers_binary_br2 %>%
  pivot_longer(
    cols = -marker,
    names_to = "cell_type",
    values_to = "value"
  ) %>% 
  mutate(marker = factor(marker,
                         levels = c("CD3D", "CD3E", "CD3G", "CCR7", "SELL", "CD8A", "CD8B", "LTB", "IL7R", "S100A4", "GZMB", "PRF1", "FCGR3A", "NKG7", "GNLY", "KLRC1", "CD14", "LYZ", "S100A9", "CSF3R", "CSF1R", "LYN", "IFITM1", "IFITM2", "IFITM3", "CD79A", "MS4A1", "CD1C", "ITGAX", "CLEC4C", "GP9", "ITGA2B", "PF4", "PPBP")),
          cell_type = factor(cell_type,
                             levels = c("CD4T", "CD8T", "NK", "mono", "B", "DC", "megak")))

```

```{r fig.dim = c(12,10)}
m1 <- ggplot(markers_long_sp, aes(x = marker, y = cell_type)) +
  geom_point(aes(size = abs(value), fill = as.factor(value)), shape = 21, color = "black") +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1))
m1
#ggsave("specific_celltype_by_markerexp_dotplot.png", plot = m1, device = png, width = 16, height = 12, unit = "cm", bg = "white", path = paste0(file_path, "marker_genes/") )

m2 <- ggplot(markers_long_br, aes(x = marker, y = cell_type)) +
  geom_point(aes(size = abs(value), fill = as.factor(value)), shape = 21, color = "black") +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1))
m2
#ggsave("broad_celltype_by_markerexp_dotplot.png", plot = m2, device = png, width = 16, height = 12, unit = "cm", bg = "white", path = paste0(file_path, "marker_genes/") )

m3 <- ggplot(markers_long_br2, aes(x = marker, y = cell_type)) +
  geom_point(aes(size = abs(value), fill = as.factor(value)), shape = 21, color = "black") +
  coord_flip() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1))
m3
#ggsave("broad_celltype_by_allmarkerexp_dotplot.png", plot = m3, device = png, width = 16, height = 12, unit = "cm", bg = "white", path = paste0(file_path, "marker_genes/") )

```

```{r fig.dim = c(12,10)}
marker_genes <- c("CD3D", "CD3E", "CD3G", "CCR7", "SELL", "CD8A", "CD8B", "LTB", "IL7R", "S100A4", "GZMB", "PRF1", "FCGR3A", "NKG7", "GNLY", "KLRC1", "CD14", "LYZ", "S100A9", "CSF3R", "CSF1R", "LYN", "IFITM1", "IFITM2", "IFITM3", "CD79A", "MS4A1", "CD1C", "ITGAX", "CLEC4C", "GP9", "ITGA2B", "PF4", "PPBP")

d1 <- DotPlot(scCHANTS_pbmc, features = marker_genes, assay = "RNA", group.by = "seurat_clusters", cols = c("lightgrey", "blue")) +
  coord_flip() +
  ggtitle("scCHANTS PBMC cluster by marker exp run1")
d1

#ggsave("clusters_by_markerexp_dotplot.pdf", plot = d1, device = pdf, width = 20, height = 18, unit = "cm", bg = "white", path = paste0(plot_path, "cell_types/") )
#ggsave("clusters_by_markerexp_dotplot.png", plot = d1, device = png, width = 20, height = 18, unit = "cm", bg = "white", path = paste0(plot_path, "cell_types/") )


marker_genes_broad <- c("CD3D", "CD3E", "CD3G", "CD8A", "CD8B", "GZMB", "PRF1", "CD14", "CD79A", "GP9", "ITGA2B", "PF4", "PPBP")


d2 <- DotPlot(scCHANTS_pbmc, features = marker_genes_broad, assay = "RNA", group.by = "seurat_clusters", cols = c("lightgrey", "blue")) +
  coord_flip() +
  ggtitle("scCHANTS PBMC cluster by marker exp run1")
d2

#ggsave("clusters_by_markerexp_dotplot_broad.pdf", plot = d2, device = pdf, width = 20, height = 14, unit = "cm", bg = "white", path = paste0(plot_path, "cell_types/") )
#ggsave("clusters_by_markerexp_dotplot_broad.png", plot = d2, device = png, width = 20, height = 14, unit = "cm", bg = "white", path = paste0(plot_path, "cell_types/") )


```

