---
title: "script_1.1_filter_data"
author: "GKM"
date: "2025-07-15"
output:
  html_document:
    keep_md: TRUE
---

This script is a test script to work with 20250616_benchmark sample to set up a pipeline:

* unzip files
* load into a seurat object
* (in future scirpts will need to do data integration)
* filter and quality control
* normalisation
* (batch effect correction)
* feature selection
* dimensionality reduction
* export Rds file


# Load packages
```{r}
library("R.utils")    # need for decompressing files
library("dplyr")
library("Seurat")
library("ggplot2")
```

# Decompress the files
```{r}
# only need to do once
# gunzip("/cephfs/volumes/hpc_data_prj/id_hill_sims_wellcda/c1947608-5b3a-4d60-8179-b8e0779d7319/scratch_tmp/scCHANTS/20250616_benchmark/20250616_benchmark/matrix.mtx.gz", temporary = TRUE, remove = FALSE)
# gunzip("/cephfs/volumes/hpc_data_prj/id_hill_sims_wellcda/c1947608-5b3a-4d60-8179-b8e0779d7319/scratch_tmp/scCHANTS/20250616_benchmark/20250616_benchmark/barcodes.tsv.gz", temporary = TRUE, remove = FALSE)
# gunzip("/cephfs/volumes/hpc_data_prj/id_hill_sims_wellcda/c1947608-5b3a-4d60-8179-b8e0779d7319/scratch_tmp/scCHANTS/20250616_benchmark/20250616_benchmark/features.tsv.gz", temporary = TRUE, remove = FALSE)

```

# Load data
```{r}
# load the unzipped files
# gene.column = 2 gives gene symbol/name instead of ENSEMBL ID
scCHANTS_data <- Read10X(data.dir = "/cephfs/volumes/hpc_data_prj/id_hill_sims_wellcda/c1947608-5b3a-4d60-8179-b8e0779d7319/scratch_tmp/scCHANTS/20250616_benchmark/20250616_benchmark/", gene.column=2)

# create seurat object
# could add arguments for min.cells or min.features to be included
scCHANTS <- CreateSeuratObject(counts = scCHANTS_data, project = "scCHANTS")

Assays(scCHANTS)
```

Rename HTO and ADT ids
```{r}
# add HTO and ADT assay to the seurat object as it is by default ignored by CreateSeuratObject
# extract feature names
HTO_names <- rownames(scCHANTS_data$`Antibody Capture`) [1:14]
ADT_names <- rownames(scCHANTS_data$`Antibody Capture`) [15:18]

# extract count matrices
HTO_counts <- scCHANTS_data$`Antibody Capture` [HTO_names, , drop=FALSE]
ADT_counts <- scCHANTS_data$`Antibody Capture` [ADT_names, , drop=FALSE]

# check names and order
rownames(HTO_counts)
rownames(ADT_counts)

# create mapping from long to short names. using - not _ as cannot have underscores in seurat object names
name_to_id <- c(
  "TotalSeq™-C0251 anti-human Hashtag 1 Antibody"  = "HTO-1",
  "TotalSeq™-C0252 anti-human Hashtag 2 Antibody"  = "HTO-2",
  "TotalSeq™-C0253 anti-human Hashtag 3 Antibody"  = "HTO-3",
  "TotalSeq™-C0254 anti-human Hashtag 4 Antibody"  = "HTO-4",
  "TotalSeq™-C0255 anti-human Hashtag 5 Antibody"  = "HTO-5",
  "TotalSeq™-C0256 anti-human Hashtag 6 Antibody"  = "HTO-6",
  "TotalSeq™-C0257 anti-human Hashtag 7 Antibody"  = "HTO-7",
  "TotalSeq™-C0258 anti-human Hashtag 8 Antibody"  = "HTO-8",
  "TotalSeq™-C0259 anti-human Hashtag 9 Antibody"  = "HTO-9",
  "TotalSeq™-C0260 anti-human Hashtag 10 Antibody" = "HTO-10",
  "TotalSeq™-C0296 anti-human Hashtag 11 Antibody" = "HTO-11",
  "TotalSeq™-C0262 anti-human Hashtag 12 Antibody" = "HTO-12",
  "TotalSeq™-C0263 anti-human Hashtag 13 Antibody" = "HTO-13",
  "TotalSeq™-C0264 anti-human Hashtag 14 Antibody" = "HTO-14",
  
  # ADTs
  "TotalSeq-C anti human CD14 (clone M5E2, cat 301859)"             = "ADT-CD14",
  "TotalSeq-C anti human CD16 (clone 3G8 cat 302065)"               = "ADT-CD16",
  "TotalSeq™-C0045 anti-human CD4 Antibody (clone SK3, cat 344651)" = "ADT-CD4",
  "TotalSeq™-C0046 anti-human CD8 Antibody (clone SK1, cat 344753)" = "ADT-CD8"
)

# rename the rows (features) using the mapping
rownames(HTO_counts) <- unname(name_to_id[rownames(HTO_counts)])
rownames(ADT_counts) <- unname(name_to_id[rownames(ADT_counts)])

# check names have changed
rownames(HTO_counts)
rownames(ADT_counts)
```

Add HTO and ADT assays to seurat object
```{r}
# add HTO assay
scCHANTS[["HTO"]] <- CreateAssayObject(counts = HTO_counts)

# add ADT assay
scCHANTS[["ADT"]] <- CreateAssayObject(counts = ADT_counts)

# validate assays were added
Assays(scCHANTS)
```


```{r}
# check default assay
DefaultAssay(scCHANTS)

# if needed, can change default assay
DefaultAssay(scCHANTS) <- "RNA"

# check number of cells (samples) and features looks right
scCHANTS
scCHANTS@assays$HTO
scCHANTS@assays$ADT

# free space
rm(scCHANTS_data)
```


# Demulitplex HTOs
For 20250616 data, there is shallow sequencing depth which is resulting in low read cutoffs (2-18, would expect >100), flattened ridgeplots (expect a clear bimodal distribution at 0 and positives not shallow enrichment curves) and (when normalised across cells) low negative calling. Also shown by low peak of total HTO counts per cell histogram
Should check in future experiments whether to normalise across features or cells (use margin argument)

https://satijalab.org/seurat/articles/hashing_vignette

HTODemux:

* Perform a k-medoid clustering on the normalized HTO values, which initially separates cells into K(# of samples)+1 clusters.
* Calculate a ‘negative’ distribution for HTO. For each HTO, the cluster with the lowest average value is used as the negative group.
* For each HTO, fit a negative binomial distribution to the negative cluster. 0.99 quantile of this distribution is used as a threshold.
* Based on these thresholds, each cell is classified as positive or negative for each HTO.
* Cells that are positive for more than one HTOs are annotated as doublets.

```{r}
DefaultAssay(scCHANTS) <- "HTO"

# normalise HTO counts, standard is centered log-ratio (CLR) transformation
# can normalise across features (margin = 1) or across cells (margin = 2)
# comment on github form satija lab that recommend normalising across tags if variation between hash performances but since have very low signals, will try across cells
scCHANTS <- NormalizeData(scCHANTS, assay = "HTO", normalization.method = "CLR", margin = 2)

# using default threshold (quantile of inferred negative distribution for each hashtag), default kfunc is clara (for fast k-medoids clustering on large applications, it is faster and more memory-efficient as it repeatedly samples subsets and clusters those)
scCHANTS <- HTODemux(scCHANTS, assay = "HTO", positive.quantile = 0.99, kfunc = "clara")
```

Visualise demultiplexing
```{r}
# look at classification
table(scCHANTS$HTO_classification.global)

# visualise enrichment for some HTOs, group cells based on the max HTO signal
Idents(scCHANTS) <- "HTO_maxID"
RidgePlot(scCHANTS, assay = "HTO", features = rownames(scCHANTS[["HTO"]])[1:4], ncol = 2)

# look at pairs to see if mutually exclusive
FeatureScatter(scCHANTS, feature1 = "HTO-1", feature2 = "HTO-2")

# compare number UMIs
Idents(scCHANTS) <- "HTO_classification.global"
VlnPlot(scCHANTS, features = "nCount_RNA", pt.size = 0.1, log = TRUE)
```

```{r}
# first remove negative cells from the object
scCHANTS_nonegs <- subset(scCHANTS, idents = "Negative", invert = TRUE)

# calculate a tSNE embedding of the HTO data
DefaultAssay(scCHANTS_nonegs) <- "HTO"

scCHANTS_nonegs <- ScaleData(scCHANTS_nonegs, features = rownames(scCHANTS_nonegs),
    verbose = FALSE)

scCHANTS_nonegs <- RunPCA(scCHANTS_nonegs, features = rownames(scCHANTS_nonegs), approx = FALSE)

scCHANTS_nonegs <- RunUMAP(scCHANTS_nonegs, dims = 1:8, perplexity = 100)

DimPlot(scCHANTS_nonegs)
Idents(scCHANTS_nonegs) <- "HTO_classification"
DimPlot(scCHANTS_nonegs)
```
Create a heatmap of HTO signals
```{r}
HTOHeatmap(scCHANTS, assay = "HTO", ncells = 5000)
```

Visualise RNA clustering
This is just mock to check for batch effects. will do clusteirng etc later on after QC
```{r}
# extract the singlets
scCHANTS_singlets <- subset(scCHANTS, idents = "Singlet")

DefaultAssay(scCHANTS_singlets) <- "RNA"

# set "counts.Gene Expression" to "counts" so cna run find vairbale features on just this assay, not antibody capture as well 
scCHANTS_singlets <- SetAssayData(object = scCHANTS_singlets, assay = "RNA", layer = "counts", new.data = GetAssayData(scCHANTS_singlets, assay = "RNA", layer = "counts.Gene Expression") )

# select the top 1000 most variable features
scCHANTS_singlets <- FindVariableFeatures(scCHANTS_singlets, assay = "RNA", selection.method = "vst", layers = "counts")

# normalise and scale RNA data, only scale the variable features here for efficiency
scCHANTS_singlets <- NormalizeData(scCHANTS_singlets, assay = "RNA", layer = "counts")
scCHANTS_singlets <- ScaleData(scCHANTS_singlets, features = VariableFeatures(scCHANTS_singlets))

# run PCA
scCHANTS_singlets <- RunPCA(scCHANTS_singlets, features = VariableFeatures(scCHANTS_singlets))

# select the top 10 PCs for clustering and tSNE based on PCElbowPlot
scCHANTS_singlets <- FindNeighbors(scCHANTS_singlets, reduction = "pca", dims = 1:10)
scCHANTS_singlets <- FindClusters(scCHANTS_singlets, resolution = 0.6, verbose = FALSE)
scCHANTS_singlets <- RunUMAP(scCHANTS_singlets, reduction = "pca", dims = 1:10)

# project singlet identities on UMAP visualisation
DimPlot(scCHANTS_singlets, group.by = "HTO_classification")

# show only a subset
subset1 <- subset(scCHANTS_singlets, subset = HTO_maxID %in% c("HTO-1", "HTO-2", "HTO-3") )
DimPlot(subset1, group.by = "HTO_classification")

subset2 <- subset(scCHANTS_singlets, subset = HTO_maxID %in% c("HTO-7", "HTO-8", "HTO-9") )
DimPlot(subset2, group.by = "HTO_classification")

rm(subset1, subset2)
```

## Check HTO data
20250616 data, shallow sequencing has lead to low HTO signal, check the signal here
```{r}
# check HTO counts
HTO_counts <- GetAssayData(scCHANTS, assay = "HTO", layer = "counts")
summary(Matrix::colSums(HTO_counts))
## shows UMI counts are not too low

## checked HTOs are labelled as antibody capture in cell ranger
# check library size
hist(Matrix::colSums(HTO_counts), breaks = 150, main = "Total HTO UMIs per Cell", xlab = "HTO UMIs")
## not peaking around 0 which is good

```





# Reassign scCHANTS to only singlets
```{r}
scCHANTS_full <- scCHANTS
scCHANTS <- subset(scCHANTS, subset = HTO_classification.global == "Singlet")

# remove to save memory
rm(ADT_counts, HTO_counts, ADT_names, HTO_names, name_to_id, scCHANTS_nonegs, scCHANTS_singlets)

```

# Add metadata
```{r}
# examine existing metadata
metadata <- scCHANTS@meta.data

dim(metadata)
head(metadata)
summary(metadata$nCount_RNA)

# read in additional sample metadata
metadata_samples <- read.csv("/cephfs/volumes/hpc_data_prj/id_hill_sims_wellcda/c1947608-5b3a-4d60-8179-b8e0779d7319/scratch_tmp/scCHANTS/20250616_benchmark/sample_metadata.csv")

head(metadata_samples)

# steps to add to seurat object

# check HTO values are all singles, no doublets
table(scCHANTS$HTO_classification) 

# align each cell's HTO id to row in the metadata_samples, rownames here represent HTOs but can't have duplicate rownames hence 4.1 etc
hto_info <- metadata_samples[match(scCHANTS$HTO_classification, metadata_samples$HTO), ]
head(hto_info)

# add each row to new vairbale in metadata
scCHANTS$sample_id <- hto_info$sample
scCHANTS$timepoint <- hto_info$timepoint
scCHANTS$treatment <- hto_info$treatment
scCHANTS$PBMC_or_T <- hto_info$PBMC_or_T

# check new metadata
metadata <- scCHANTS@meta.data

dim(metadata)
colnames(metadata)
metadata[14:17, 14:17]
```

# Separate PBMC and sorted T cell samples
```{r}
DefaultAssay(scCHANTS) <- "RNA"

scCHANTS_pbmc <- subset(scCHANTS, subset = PBMC_or_T == "PBMC")
DefaultAssay(scCHANTS_pbmc) <- "RNA"

scCHANTS_t <- subset(scCHANTS, subset = PBMC_or_T == "T")
DefaultAssay(scCHANTS_t) <- "RNA"
```

# PBMC pre-processing

## Quality control on PBMCs
```{r}
# calculate mt contamination and add column to metadata
scCHANTS_pbmc[["percent_mt"]] <- PercentageFeatureSet(scCHANTS_pbmc, pattern = "^MT-")

# calculate mt contamination and add column to metadata
scCHANTS_pbmc[["percent_rb"]] <- PercentageFeatureSet(scCHANTS_pbmc, pattern = "^RP[SL]")

# plot on violin plot
VlnPlot(scCHANTS_pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent_mt", "percent_rb"), ncol = 2, alpha = 0.3)
## before filtering if mt counts are at 0 check you're in RNA assay as default, check head(rownames(scCHANTS_pbmc)) shows gene names


# feature scatter
p1 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "timepoint") +
  theme(legend.position="none")
p2 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_mt", group.by = "timepoint") + 
  theme(legend.position="none")
p3 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_rb", group.by = "timepoint") +
  theme(legend.position="none")
p4 <- FeatureScatter(scCHANTS_pbmc, feature1 = "percent_mt", feature2 = "percent_rb", group.by = "timepoint") +
  theme(legend.position="none")
(p1 + p2) / (p3 + p4)

```

Do subsetting
```{r}
# can subset straightaway based on QC metrics e.g. :
#scCHANTS_pbmc <- subset(scCHANTS_pbmc, subset = nFeature_RNA > 200  & nFeature_RNA < 2500 & percent_mt < 5) 

# or assign new metadata column
# start with all cells labeled as "initial"
scCHANTS_pbmc$QC <- "initial"

# label low nFeature_RNA (< 150)
## i.e. if nFeature_RNA < 150 (and previously passed QC as cautionary step), call it low_nFeature, otherwise leave as initial

scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$nFeature_RNA < 150 & scCHANTS_pbmc$QC == "initial",
  "low_nFeature",
  scCHANTS_pbmc$QC
)

# label high nFeature_RNA (> 2000)
scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$nFeature_RNA > 2000 & scCHANTS_pbmc$QC == "initial",
  "high_nFeature",
  scCHANTS_pbmc$QC
)

# label high mitochondrial content (> 5%)
scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$percent_mt > 5 & scCHANTS_pbmc$QC == "initial",
  "high_mt",
  scCHANTS_pbmc$QC
)

# label combined: low_nFeature + high_mt
scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$nFeature_RNA < 150 & scCHANTS_pbmc$percent_mt > 5 & scCHANTS_pbmc$QC != "pass",
  "low_nFeature_high_mt",
  scCHANTS_pbmc$QC
)

# label combined: high_nFeature + high_mt
scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$nFeature_RNA > 2000 & scCHANTS_pbmc$percent_mt > 5 & scCHANTS_pbmc$QC != "pass",
  "high_nFeature_high_mt",
  scCHANTS_pbmc$QC
)

# finally, label cells that pass all filters
scCHANTS_pbmc$QC <- ifelse(
  scCHANTS_pbmc$nFeature_RNA >= 150 &
    scCHANTS_pbmc$nFeature_RNA <= 2000 &
    scCHANTS_pbmc$percent_mt <= 5 &
    scCHANTS_pbmc$QC == "initial",
  "pass",
  scCHANTS_pbmc$QC
)

```

Replot QC plots, separated by QC result
```{r}
# plot on violin plot
VlnPlot(scCHANTS_pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent_mt", "percent_rb"), ncol = 2, alpha = 0.3, , group.by = "QC")

# feature scatter, coloured by QC overlaid
p1 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", split.by = "timepoint", group.by = "QC") +
  theme(legend.position="none")
p2 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_mt", split.by = "timepoint", group.by = "QC") + 
  theme(legend.position="none")
p3 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_rb", split.by = "timepoint", group.by = "QC") +
  theme(legend.position="none")
p4 <- FeatureScatter(scCHANTS_pbmc, feature1 = "percent_mt", feature2 = "percent_rb", split.by = "timepoint", group.by = "QC") +
  theme(legend.position="none")
(p1 + p2) / (p3 + p4)


# feature scatter, coloured by QC split
p1 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", split.by = "QC", group.by = "QC") +
  theme(legend.position="none")
p2 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_mt", split.by = "QC", group.by = "QC") + 
  theme(legend.position="none")
p3 <- FeatureScatter(scCHANTS_pbmc, feature1 = "nCount_RNA", feature2 = "percent_rb", split.by = "QC", group.by = "QC") +
  theme(legend.position="none")
p4 <- FeatureScatter(scCHANTS_pbmc, feature1 = "percent_mt", feature2 = "percent_rb", split.by = "QC", group.by = "QC") +
  theme(legend.position="none")
(p1 + p2) / (p3 + p4)

scCHANTS_pbmc <- subset(scCHANTS_pbmc, subset = QC == "pass")
```

## Normalisaiton of PBMCs
Default is lognormalise, a global-scaling normalisation method normalises feature expression measurements for each cell by total expression, x by 10,000 by default and log-transforms the result. It stores these in [["RNA"]]$data
Assumes each cell originally contains same number of RNA molecules
Can also use SCTransform (also replaces FindVariableFeatures and ScaleData)
```{r}
scCHANTS_pbmc <- NormalizeData(scCHANTS_pbmc, normalization.method = "LogNormalize", scale.factor = 10000)

# find variable features
scCHANTS_pbmc <- FindVariableFeatures(scCHANTS_pbmc, selection.method = "vst", nfeatures = 2000)

# identify 10 most highly variable genes
top10_features <- head(VariableFeatures(scCHANTS_pbmc), 10)

# to plot these 10 most variable genes
p1 <- VariableFeaturePlot(scCHANTS_pbmc)
p2 <- VariableFeaturePlot(scCHANTS_pbmc) %>% LabelPoints(points = top10_features, repel = TRUE)
p1 + p2

# scale data
all.genes <- rownames(scCHANTS_pbmc)
scCHANTS_pbmc <- ScaleData(scCHANTS_pbmc, features = all.genes)
```

## Dim red on PBMC
```{r}
# run PCA, variable features already calculated and in the saved Rds file
scCHANTS_pbmc <- RunPCA(scCHANTS_pbmc, features = VariableFeatures(object = scCHANTS_pbmc))

# visualise
VizDimLoadings(scCHANTS_pbmc, dims = 1:2, reduction = "pca")

DimHeatmap(scCHANTS_pbmc, dims = 1, cells = 500, balanced = TRUE)

DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)

# generate elbow plot and save as a variable
elbow_plot <- ElbowPlot(scCHANTS_pbmc) +
  labs(title = paste(scCHANTS_pbmc_name, "elbowplot", sep = "_"))
## decide on pc number

# clustering
scCHANTS_pbmc <- FindNeighbors(scCHANTS_pbmc, dims = 1:10) 

scCHANTS_pbmc <- FindClusters(scCHANTS_pbmc, resolution = 1.0)

scCHANTS_pbmc <- RunUMAP(scCHANTS_pbmc, dims = 1:10)

# plot umap coloured by cell_type and cell_type_lowerres
DimPlot(scCHANTS_pbmc, reduction = "umap", group.by = "sample_id")
DimPlot(scCHANTS_pbmc, reduction = "umap", group.by = "timepoint")

```

# save PBMC pre-processed RDS (takes a long time)
```{r}
#saveRDS(object=scCHANTS_pbmc, file ="/scratch_tmp/prj/id_hill_sims_wellcda/1M-scBloodNL/scCHANTS_pbmc_processed.Rds")

```


# T cell pre-processing
## Quality control on T cells

